Human judgement forecasting of COVID-19 in the UK

### 2024-02-01 Analysis.rmd
Update Figure 1 and correct some typos

```{r}
plot_data_simple <- function(weekly_truth, daily_truth) {
  weekly_truth |>
    mutate(status = str_to_title(status)) |>
  ggplot(aes(x = target_end_date, y = true_value)) +
  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
           fill = "darkolivegreen3", alpha = .1) + 
  geom_line() + 
  geom_point(size = 0.8) + 
  geom_bar(data = daily_truth |>
             mutate(true_value = true_value * 7), 
           stat = "identity", 
           fill = "skyblue3", alpha = 0.4) + 
  facet_wrap(~ target_type, 
             nrow = 2, scale = "free_y") + 
  scale_y_continuous(labels = label_fn) +
  theme_scoringutils() +
  labs(y = "Observed values", x = "Date", color = "Data status")
}

p_data <- plot_data_simple(weekly_truth, daily_truth)
ggsave("output/figures/overview-data.png", plot = p_data, 
       width = 7, height = 5)

>>>>

```{r}
plot_data_simple <- function(weekly_truth, daily_truth) {
  weekly_truth |>
    mutate(status = str_to_title(status)) |>
  ggplot(aes(x = target_end_date, y = true_value, colour = "Weekly data")) +
  annotate("rect", xmin = time_start, xmax = time_stop, ymin = -Inf, ymax = Inf,
           fill = "darkolivegreen3", alpha = .1) + 
  geom_line() + 
  geom_point(size = 0.8) + 
  geom_bar(data = daily_truth |>
             mutate(true_value = true_value * 7), 
           aes(fill = "7x daily cases"), 
           colour = NA,
           stat = "identity",
           alpha = 0.4) + 
  facet_wrap(~ target_type, 
             nrow = 2, scale = "free_y") + 
  scale_y_continuous(labels = label_fn) +
  scale_fill_manual(values = c("7x daily cases" = "skyblue3")) +
  scale_color_manual(values = c("Weekly data" = "black")) +
  theme_scoringutils() +
  labs(y = "Observed values", x = "Date", color = " ", fill = " ")
}

(p_data <- plot_data_simple(weekly_truth, daily_truth))
ggsave("output/figures/overview-data.png", plot = p_data, 
       width = 7, height = 5)

### 2024-02-01 manuscript/manuscript.tex
Update Figure 1 and correct some typos

%% Default: numerical citations
\usepackage[numbers]{natbib}
\bibliographystyle{custom}

@@

Ground-truth data on daily reported test positive cases and deaths linked to COVID-19 were provided by the European Forecast Hub and sourced from the Johns Hopkins University (JHU). Data were subject to reporting artifacts and revisions. All data points were marked as anomalous retrospectively by the European Forecast Hub if in subsequent updates data was changed by more than 5 percent. In August 2022 JHU switched the data source for their UK death numbers from "deaths within 28 days of a positive COVID test" to "Deaths with COVID-19 on the death certificate" and revised all their past data to guarantee consistency. The 2021 UK ground truth death data as it was made available through the European Forecast Hub in 2021 is therefore substantially different and on average lower than the data available as of early 2023. The data and revisions are displayed in Figure 2 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}. All results presented here were derived based on the original data available in 2021, which were available through the European COVID-19 Forecast Hub GitHub repository (\url{https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe}).

@@

@@

The median number of unique participants in any given week was 17, the minimum was 6 and the maximum was 51. This was higher than the number of participants in \cite{bosseComparingHumanModelbased2022} (which had a median number of 6, a minimum of 2, and a maximum 10). With respect to the number of submissions from an individual participant, we observed similar patterns as \cite{bosseComparingHumanModelbased2022}: An individual forecaster participated on average in 2.6 weeks out of 13. The median number of submissions from a single individual was one, meaning that similar to \citep{bosseComparingHumanModelbased2022} most forecasters dropped out after their first submission. Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study (S. Abbott). The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}A). The median number of "non-experts" (11 for cases, 10 for deaths) was higher than the median number of "experts" (8 for cases and deaths) (see Figure \ref{fig:num-forecasters}B).

@@

\input performance-table.tex

@@

@@

@@

\input{manuscript/performance-table-experts}

@@

@@

In their original study, conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths. They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators. \citet{mcandrewChimericForecastingCombining2022} similarly found in their study that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19. 
Results in our study do not directly support this pattern, but given the low number of observations also do not provide strong evidence against it. In this study, the combined crowd ensemble performed better than the Hub ensemble on both cases and deaths on the log scale, and worse on the natural scale. Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths. 
During the study period, the case fatality ratio (CFR) changed quite quickly. This was related to the rise of the Delta variant in the UK in the beginning of May with its higher case fatality rate \citep{bastIncreasedRiskHospitalisation2021}. In addition, the age distribution of cases changed over the course of the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}.
One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update . A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study used a mean ensemble.

@@

This study explored a novel method of forecasting infectious diseases that combines a human forecast of the estimated effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths. One appeal of this approach is that the forecaster can directly forecast the generative process and how they believe it is affected by interventions and changes in behaviour. Computational modelling then takes care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, and the relationship between different indicators. This could help reduce cognitive load, and make it easier to synthesise various sources in information into a single forecast, at least for forecasters who have an intuitive understanding of $R_t$. Though we note all of these modelling steps and the construction of the model itself requires the human constructing the model to make assumptions. Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience. In our study, $R_t$ forecasts of cases were comparable to direct forecasts, with a tendency for good forecasts to improve when using the $R_t$ method and worse forecasts to deteriorate even more. Sample sizes, however, were very low. 
Given that forecasters could simulate cases in the app, it is also possible that forecasters were in fact directly forecasting cases. 
$R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths. The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and was assumed to be constant over the four week forecast horizon. However, as mentioned before, the CFR evolved during the study period. Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast and also had no way to adjust the CFR manually, likely impacting forecast accuracy. Giving human forecasters the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model. Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon though this approach may still have struggled to cope with the rapid changes observed during the study period. Another important limitation is that we didn't have full sample trajectories of the $R_t$-values predicted by forecasters. Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons, which likely negatively affected forecasts. One potential way to disentangle the effect of the convolution model from the $R_t$ forecasts would have been to use the human forecasts for cases as an input to the second computational model, which could then have simulated deaths. Future work could expose forecasters to different combinations of these options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.

@@

>>>>

%% Default: numerical citations
\usepackage[numbers]{natbib}
% \bibliographystyle{custom}

@@

Ground-truth data on daily reported test positive cases and deaths linked to COVID-19 were provided by the European Forecast Hub and sourced from the Johns Hopkins University (JHU). Data were subject to reporting artifacts and revisions. All data points were marked as anomalous retrospectively by the European Forecast Hub if in subsequent updates data was changed by more than 5 percent. In August 2022 JHU switched the data source for their UK death numbers from "deaths within 28 days of a positive COVID test" to "Deaths with COVID-19 on the death certificate" and revised all their past data to guarantee consistency. The 2021 UK ground truth death data as it was made available through the European Forecast Hub in 2021 is therefore substantially different and on average lower than the data available as of early 2023. Data revisions are displayed in Figure 2 in the Supplementary Information \cite{bosseSupplementaryInformationHuman2023}. All results presented here were derived based on the original data available in 2021, which were available through the European COVID-19 Forecast Hub GitHub repository (\url{https://github.com/covid19-forecast-hub-europe/covid19-forecast-hub-europe}).

@@

\subsection*{Observed values}

The study period (forecasts were made between May 24 and August 16, 2021, for targets between May 29 and September 11, 2021) was characterised by an increase in the number of cases and deaths in the United Kingdom. Reported cases in particular rose rapidly compared to pre-study levels, with a peak on July 17, 2021, followed by a trough and another subsequent increase in numbers. Death numbers remained almost constant in the first four weeks of the study period, followed by a steady increase until the end of the study period in September 2021. This increase in the case and death numbers coincides with the rise of the Delta variant in the UK at the beginning of May \citep{bastIncreasedRiskHospitalisation2021} as well as the  European Football Championship \citep{dehningImpactEuro20202023}. An overview of the reported case and death numbers is shown in Figure \ref{fig:overview-data}.

\begin{figure*}
\centering
\includegraphics[width=0.99\textwidth]{../output/figures/overview-data.png}
\caption{\bf{Observed cases and deaths of COVID-19 in the UK}. A: Observed daily (bars) and weekly (black lines and points) numbers of cases and deaths as available through the European Forecast Hub when the study concluded in 2021. Daily numbers were multiplied by seven in order to appear on the same scale as weekly numbers.}
\label{fig:overview-data}
\end{figure*}

@@

The median number of unique participants in any given week was 17, the minimum was 6 and the maximum was 51. This was higher than the number of participants in \cite{bosseComparingHumanModelbased2022} (which had a median number of 6, a minimum of 2, and a maximum 10). With respect to the number of submissions from an individual participant, we observed similar patterns as \cite{bosseComparingHumanModelbased2022}: An individual forecaster participated on average in 2.6 weeks out of 13. The median number of submissions from a single individual was one, meaning that similar to \citep{bosseComparingHumanModelbased2022} most forecasters dropped out after their first submission. Only five participants submitted a forecast in ten or more weeks and only two submitted a forecast in all thirteen weeks, one of whom is an author on this study (S. Abbott). Three other authors participated in the study (S. Funk, N. Bosse, and E. van Leuwwen). A total of 535 forecasts were submitted by human forecasters, 118 (22\%) of these were submitted by authors of this study. The number of direct forecasts (median: 13 for cases and 12 for deaths) was higher than the number of $R_t$ forecasts (median: 6 for both cases and deaths) in all weeks (see Figure \ref{fig:num-forecasters}A). The median number of "non-experts" (11 for cases, 10 for deaths) was higher than the median number of "experts" (8 for cases and deaths) (see Figure \ref{fig:num-forecasters}B).

@@

% \input performance-table.tex

@@

At the beginning of the study period, human forecasters as well as the Forecast Hub ensemble, consistently underpredicted case numbers (see Figure \ref{fig:forecasts-scores}A). All forecasting approaches overshot the peak in case numbers on July 17, 2021, overpredicting case numbers severely in the three weeks after, followed again by a small tendency to underpredict when case numbers rose once more in the 4th week after the peak.

@@

In the first part of the study period, most forecasting approaches (albeit not the direct crowd forecasts), showed a tendency to overpredict the increase in death numbers (see Figure \ref{fig:forecasts-scores}B). All forecasting approaches started to underpredict death numbers four weeks after the peak in case numbers on July 17, 2021, expecting a consequent drop in deaths that did not occur.

@@

\input{performance-table-experts}

@@

% Study period. 
The study period was marked by an increase in both case and death numbers. Case numbers rose quickly compared to the pre-study period, peaking on July 17, 2021, followed by a trough and a subsequent further increase. Forecasts displayed a familiar pattern, where forecasters tended to underpredict while case numbers were rising, and overpredict while case numbers were falling, particularly following a peak. Death numbers were increasing more slowly than during the previous peak in January 2021, in line with the beginning of vaccination efforts and a greater natural immunity in the population. The peak in case numbers did not correspond to a peak in death numbers, suggesting some decoupling of case and death numbers such as would be expected from effects of immunity that are stronger in preventing severe disease than any symptoms. Forecasters tended to overpredict death numbers in the beginning, while underpredicting them in the end, expecting death numbers to fall after the peak in cases. The study period coincides with the rise of the Delta variant in the UK \citep{bastIncreasedRiskHospitalisation2021}, as well as the 2021 European Football Championship, which likely shifted the age distribution towards younger cases \citep{dehningImpactEuro20202023}.

@@

In their original study, conducted in Germany and Poland, \citet{bosseComparingHumanModelbased2022} found that humans outperformed an ensemble of computational models when predicting cases, but not when predicting deaths. They hypothesised that computational models might have an advantage over human forecasters when predicting deaths, benefiting from the ability to model the delays and epidemiological relationships between different leading and lagged indicators. \citet{mcandrewChimericForecastingCombining2022} similarly found in their study that humans performed comparably to an ensemble of computational models for cases, but not for predictions of deaths of COVID-19. 
Results in our study do not directly support this pattern, but given the low number of observations also do not provide strong evidence against it. In this study, the combined crowd ensemble performed better than the Hub ensemble on both cases and deaths on the log scale, and worse on the natural scale. Direct forecasts, which would be most comparable to the forecasts in \citep{bosseComparingHumanModelbased2022}, performed worse than the Hub ensemble on cases and better on deaths. 
During the study period, the case fatality ratio (CFR) likely changed quite quickly compared to the pre-study period. On the one hand, the rise of the Delta variant in the UK, which was first discovered in the UK in March 2021, was estimated to have a higher CFR than previous variants \citep{bastIncreasedRiskHospitalisation2021, twohigHospitalAdmissionEmergency2022}. On the other hand, the ongoing COVID-19 vaccination and growing natural immunity in the population had decreasing effects on the CFR. In addition, the age distribution of cases changed throughout the study period in Summer 2021, in parts related to the European Football Championship \citep{dehningImpactEuro20202023}. Overall, the CFR was lower than during previous peaks of COVID-19.
One possible hypothesis for the relatively good performance of human forecasts for deaths compared to previous studies might be that some models submitted to the Forecast Hub may have been more negatively affected by the changes in CFR during the study period than human forecasters or have been slower to update. The present study only saw a steady increase in death numbers, which one could argue is relatively easy to predict, making it difficult to compare forecast performance with performance in other settings. A confounding factor, when comparing results from this study and the one in Germany and Poland directly, is that we used a median ensemble to combine individual forecasts here, while the earlier study used a mean ensemble.

@@

This study explored a novel method of forecasting infectious diseases that combines a human forecast of the estimated effective reproduction number $R_t$ with epidemiological modelling to map the $R_t$ forecast to a forecast of cases and deaths. One appeal of this approach is that the forecaster can directly forecast the generative process and how they believe it is affected by interventions and changes in behaviour. Computational modelling then takes care of dealing with details such as reporting delays, generation intervals, day of the week periodicity, and the relationship between different indicators. This could help reduce cognitive load, and make it easier to synthesise various sources in information into a single forecast, at least for forecasters who have an intuitive understanding of $R_t$. Though we note all of these modelling steps and the construction of the model itself requires the human constructing the model to make assumptions. Anecdotally, forecasters familiar to the authors reported high satisfaction with the forecasting experience. In our study, $R_t$ forecasts of cases were comparable to direct forecasts, with a tendency for good forecasts to improve when using the $R_t$ method and worse forecasts to deteriorate even more. Sample sizes, however, were very low. 
Given that forecasters could simulate cases in the app, it is also possible that forecasters were in fact directly forecasting cases. 
$R_t$ forecasts of deaths (which forecasters could not see in the app) were noticeably worse than direct forecasts of deaths. The computational model underlying our $R_t$ forecasts of deaths estimated a constant CFR and delay distribution using the last 4 weeks of data, therefore updating relatively slowly to new circumstances and was assumed to be constant over the four week forecast horizon. However, as mentioned before, the CFR likely evolved during the study period. Forecasters had no way of inspecting the death forecast implied by their $R_t$ forecast, likely impacting forecast accuracy. They also had no way to adjust the CFR manually. Giving human forecasters to see their implied death forecasts, as well as the ability to adjust the CFR and other model parameters would have increased complexity of the interface, but would have solved issues with the assumptions of the underlying model. Alternatively, a more complex model could have been used which allowed for time-varying CFR estimates and forecast these changes over the forecast horizon though this approach may still have struggled to cope with the rapid changes observed during the study period. Another important limitation is that we didn't have full sample trajectories of the $R_t$-values predicted by forecasters. Rather, trajectories had to be constructed based on the distributions provided for the different forecast horizons, which likely negatively affected forecasts. One potential way to disentangle the effect of the convolution model from the $R_t$ forecasts would have been to use the human forecasts for cases as an input to the second computational model, which could then have simulated deaths. Future work could expose forecasters to different combinations of these options with the aim of separating effects of the user interface from ones related to the structure of the underlying computational model.

@@

It is plausible to hypothesise that the default baseline shown to forecasters in the app may influence their predictions. One could also interpret the $R_t$-forecast as a way of showing a different baseline forecast to the forecaster compared to the direct forecast. In our study, the default was a naive forecast with the median equal to the last value and uncertainty equal to the standard deviation of the last four changes in weekly log values. \cite{bosseComparingHumanModelbased2022} did not fnd conclusive evidence to that effect, but also did not analyse the question in detail. We suggest further research be done into potential priming effects that a default forecast can have on users.
